
\cleardoublepage
\thispagestyle{empty}

\null\vfill

{\large\noindent\emph{`We are what we think. All that we are arises with our thoughts. \\ With our thoughts, we make the world.'}}

\vspace{3\baselineskip}

\makebox[\textwidth][r]{
    \begin{minipage}{0.5\textwidth}
    Siddharta Gautama, the Buddha, \\
    Dhp I:1 as rendered by Thomas Byrom
    \end{minipage}
}
    
\vfill\null\vfill
    
\makebox[\textwidth][r]{
    \begin{minipage}[t][10\baselineskip][t]{0.5\textwidth}
    Byrom, T. (1976). \textit{Dhammapada: The Sayings of the Buddha}. Boston, MA, US: Shambhala.\nocite{byrom1976dhammapada}
    \end{minipage}
}

%%% %%% %%%


\cleardoublepage
\thispagestyle{empty}

\null\vfill

{\large\noindent\emph{`We are governed, our minds are molded, our tastes formed, our ideas suggested, largely by men we have never heard of. .~.~. we are dominated by the relatively small number of persons .~.~. who understand the mental processes and social patterns of the masses.'}}

\vspace{\baselineskip}

\makebox[\textwidth][r]{
    \begin{minipage}{0.5\textwidth}
    Edward L. Bernays \\ \null
    \end{minipage}
}
    
\vfill\null\vfill
    
\makebox[\textwidth][r]{
    \begin{minipage}[t][10\baselineskip][t]{0.5\textwidth}
    Bernays, E. L. (1928). \textit{Propaganda}. New York, NY, US: Horace Liveright.\nocite{bernays1928propaganda}
    \end{minipage}
}


%%% %%% %%%


\cleardoublepage
\thispagestyle{empty}

\null\vfill

{\large\noindent\emph{`I am Locutus of Borg. You will respond to my questions.'}}\label{quote:locutus}

\vspace{4\baselineskip}

\makebox[\textwidth][r]{
    \begin{minipage}{0.5\textwidth}
    Captain Jean-Luc Picard \\ \null
    \end{minipage}
}
    
\vfill\null\vfill
    
\makebox[\textwidth][r]{
    \begin{minipage}[t][9.5\baselineskip][t]{0.5\textwidth}
    Roddenberry, G., Moore, R.D., \& Taylor, J. (Writers), Berman, R.K. (Producer), \& Singer, A. (Director). (1993). Descent: Part I. In \textit{Star Trek: The Next Generation}. Los Angeles, CA, USA: Paramount Studios.
    \end{minipage}
}
