
\chapter*{List of Publications}
\addcontentsline{toc}{chapter}{List of Publications}

The numbered chapters in this dissertation correspond to the following published or submitted manuscripts. Included are either postprint (Chapters 1--5) or preprint (Chapter 6) versions.

\begin{enumerate}
    \setcounter{enumi}{0}
    \item Krol, L. R., Andreessen, L. M., \& Zander, T. O. (2018). Passive Brain-Computer Interfaces: A Perspective on Increased Interactivity. In C. S. Nam, A. Nijholt, \& F. Lotte (Eds.), \textit{Brain-Computer Interfaces Handbook: Technological and Theoretical Advances} (pp. 69-86). Boca Raton, FL, USA: CRC Press. ISBN: 9781498773430. (Postprint.)
    \item Krol, L. R., Haselager, P., \& Zander, T. O. (2020). Cognitive and affective probing: a tutorial and review of active learning for neuroadaptive technology. \textit{Journal of Neural Engineering, 17}(1), 012001. doi: \href{https://dx.doi.org/10.1088/1741-2552/ab5bb5}{10.1088/1741-2552/ab5bb5}.  (Postprint.)
    \item Krol, L. R., Pawlitzki, J., Lotte, F., Gramann, K., \& Zander, T. O. (2018). SEREEGA: Simulating event-related EEG activity. \textit{Journal of Neuroscience Methods, 309}, 13-24. doi: \href{https://dx.doi.org/10.1016/j.jneumeth.2018.08.001}{10.1016/j.jneumeth.2018.08.001}. (Postprint.)
    \item Krol, L. R., Mousavi, M., de Sa, V. R., \& Zander, T. O. (2018). Towards Classifier Visualisation in 3D Source Space. In \textit{2018 IEEE International Conference on Systems, Man and Cybernetics (SMC)} (pp. 71–76). doi: \href{https://dx.doi.org/10.1109/SMC.2018.00022}{10.1109/SMC.2018.00022}.  (Postprint.)
    \item Zander, T. O., Krol, L. R.\footnote{Shared first authorship.}, Birbaumer, N. P., \& Gramann, K. (2016). Neuroadaptive technology enables implicit cursor control based on medial prefrontal cortex activity. \textit{Proceedings of the National Academy of Sciences, 113}(52), 14898–14903. doi: \href{https://dx.doi.org/10.1073/pnas.1605155114}{10.1073/pnas.1605155114}.  (Postprint.)
    \item Krol, L. R., Pawlitzki, J., Gramann, K. \& Zander, T. O. (submitted). Investigating the separation of salience and valence in implicit cursor control. \textit{Proceedings of the National Academy of Sciences}. (Preprint.)
\end{enumerate}
